Revealing Transient Strain in Geodetic Data with the Radial Basis Function Finite Difference Method

### 2017-05-29 HinesHetlandGPR.tex
added more

In this section, we describe how GNSS displacement observations are used to identify transient crustal strain rates, which we denote as $\dot{\bm{\varepsilon}}(x,t)$. We consider $\dot{\bm{\varepsilon}}$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\bm{\varepsilon}}$ by first identifying transient displacements, $\bm{u}(x,t)$, which we then spatially and temporally differentiate.  As we will show in Section X, estimates of $\dot{\bm{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\bm{u}$ or $\dot{\bm{u}}$.  We make a prior assumption that each component of $\mathbf{u}$ is a three-dimensional (two spatial dimensions and time) Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(x,t) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(x,t;x',t')$ is a covariance function indicating how we expect $u_i(x,t)$ to covary with $u_i(x',t')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We then drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\bm{u}$. We further assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}(x,t;x',t') = X(x,x')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in the next section.

\subsection{Spatial and temporal covariance functions}
In this section, we list several covariance functions which are considered in this paper or could be of use in future studies.

@@

One last covariance function which is worth mentioning, even though we do not use it in this paper, is the periodic covariance function \citep{Mackay1998}, 
\begin{equation}\label{eq:Periodic}
C(x,x') = \alpha^2 \exp\left(\frac{-\sin(\frac{\pi}{\lambda} |x - x'|)^2}{2 \beta^2} \right), \ \ \ x,x' \in \mathbb{R}^1.
\end{equation}
The periodic covariance function describes an isotropic Gaussian process with realizations that have period $\lambda$.  The length-scale of features within each period is controlled by $\beta$. In terms of GNSS data, a periodic Gaussian process can be used to describe seasonal deformation. The more common method of treating seasonal deformation is to model it as a linear combination of annual and semi-annual sinusoids \citep[e.g.,][]{Blewitt2002}. We believe there is potential to use periodic Gaussian processes to explore the statistical properties of seasonal signals; however, that is outside the scope of this paper.

@@

\subsection{Constraining displacements and strain with GNSS data}
We constrain $u$ with GNSS data, which records $u$ as well as other physical and non-physical processes which we are not interested in. We describe GNSS data, $\bm{d}_*$, as the realization of a random vector $\bm{d} = [d_{ij}]^T$, where $d_{ij}$ describes observations at position $x_i$ and time $t_j$. We assume that the observations can be described as   
\begin{align}\label{eq:Data}
d_{ij} = &u(x_i,t_j) + \epsilon(x_i,t_j) + a_{i} + b_{i} t_j + \\
         &c_{1i}\sin(2 \pi t_j) + c_{2i}\cos(2 \pi t_j) + c_{3i}\sin(4 \pi t_j) + c_{4i}\cos(4 \pi t_j), 
\end{align}
where $a_i$ is an offset that is unique for each GNSS monument, $b_i$ is the steady rate of tectonic deformation at $x_i$, the sinusoids describe seasonal deformation (using units of years for $t$), and $\epsilon$ is a Gaussian process for noise that cannot be described parametrically. $\epsilon$ can consist of white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}.  For now, we describe $\epsilon$ generally as $\epsilon \sim \mathcal{N}(0,C_d)$. We consider the coefficients $a_{i}$, $b_{i}$, and $c_{kj}$, to be unknown and uncorrelated random variables. Of course, the tectonic deformation, $b_i$, is in reality spatially correlated and we could invoke tectonic model to form a prior on $b_i$. However, in our application to Cascadia, we will be using displacement time series which are long enough to sufficiently constrain $b_i$ for each station and there is no need to incorporate a prior. Likewise, the seasonal coefficients may be spatially correlated as suggested by Langbein2008?. It may be worth exploring and exploiting such a correlation in a future study.

>>>>

In this section, we describe how GNSS displacement observations are used to identify transient crustal strain rates, which we denote as $\dot{\bm{\varepsilon}}(x,t)$. We consider $\dot{\bm{\varepsilon}}$ to be spatially and temporally coherent deviations from the steady rate of strain accumulation from plate tectonics. We determine $\dot{\bm{\varepsilon}}$ by first identifying transient displacements, $\bm{u}(x,t)$, which we then spatially and temporally differentiate.  As we will show in Section X, estimates of $\dot{\bm{\varepsilon}}$ turn out to be more effective at illuminating geophysical signal than estimates of $\bm{u}$ or $\dot{\bm{u}}$.  We make a prior assumption that each component of $\mathbf{u}$ is a three-dimensional (two spatial dimensions and time) Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(x,t) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}\left((x,t),(x',t')\right)$ is a covariance function indicating how we expect $u_i(x,t)$ to covary with $u_i(x',t')$. For simplicity, we treat each component of displacement independently and ignore any potential covariance. We then drop the component subscripts with the understanding that the same analysis is being repeated to estimate the east, north, and vertical components of $\bm{u}$. We further assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((x,t),(x',t')\right) = X(x,x')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe (e.g. postseismic deformation or deformation from slow slip events), and we discuss this matter in the next section.

We constrain $u$ with GNSS data, which records $u$ as well as other physical and non-physical processes which we are not interested in. We describe GNSS observations at position $x_i$ and time $t_j$ as a realization of the random variable 
\begin{align}\label{eq:Data}
d_{ij} = &u(x_i,t_j) + \epsilon(x_i,t_j) + m_{1i} + m_{2i} t_j + \\
         &m_{3i}\sin(2 \pi t_j) + m_{4i}\cos(2 \pi t_j) + m_{5i}\sin(4 \pi t_j) + m_{6i}\cos(4 \pi t_j), 
\end{align}
where $m_1i$ is an offset that is unique for each GNSS monument, $m_2i$ is the steady rate of tectonic deformation at $x_i$, the sinusoids describe seasonal deformation (using units of years for $t_j$), and $\epsilon$ is a Gaussian process for noise that cannot be described parametrically. $\epsilon$ can consist of white noise, temporally correlated noise describing benchmark wobble \citep[e.g.,][]{Wyatt1982,Wyatt1989}, or spatially correlated noise describing common mode error \citep[e.g.,][]{Wdowinski1997}.  For now, we describe $\epsilon$ generally as $\epsilon \sim \mathcal{N}(0,C_\epsilon)$. We consider the coefficients $m_{kj}$, to be uncorrelated random variables with a diffuse prior. Of course, the tectonic deformation, $b_i$, is in reality spatially correlated and we could invoke tectonic model to form a prior on $b_i$. However, in our application to Cascadia, we will be using displacement time series which are long enough to sufficiently constrain $b_i$ for each station and there is no need to incorporate a prior. Likewise, the seasonal coefficients may be spatially correlated as suggested by Langbein2008?. It may be worth exploring and exploiting such a correlation in a future study.

\subsection{Spatial and temporal covariance functions}
We must choose an appropriate Gaussian process describing transient displacements $u$ and GNSS noise $\epsilon$. We assume that $u$ and $\epsilon$ both have zero mean, which leaves for use to select a parameterization for the covariance functions $X$, $T$, and $C_d$.  In this section, we list the covariance functions which are considered in this paper and those which could be of use in future studies. In the following discussion $\alpha$ and $\beta$ are hyperparameters. There are numerous strategies for selecting these hyperparameters (see \citet{Cressie1992}) and we save a discussion on choosing the hyperparameters for Section X.

@@

One last covariance function which is worth mentioning, even though we do not use it in this paper, is the periodic covariance function \citep{Mackay1998}, 
\begin{equation}\label{eq:Periodic}
C(x,x') = \alpha^2 \exp\left(\frac{-\sin(\frac{\pi}{\lambda} |x - x'|)^2}{2 \beta^2} \right), \ \ \ x,x' \in \mathbb{R}^1.
\end{equation}
The periodic covariance function describes an isotropic Gaussian process with realizations that have period $\lambda$.  The length-scale of features within each period is controlled by $\beta$. In terms of GNSS data, a periodic Gaussian process can be used to describe seasonal deformation. The more common method of treating seasonal deformation is to model it as a linear combination of annual and semi-annual sinusoids \citep[e.g.,][]{Dong2002}. We believe there is potential to use periodic Gaussian processes to explore the statistical properties of seasonal signals; however, that is outside the scope of this paper.

@@

\subsection{Conditioning}
Assuming that a covariance function and corresponding hyperparameters for $X$, $T$, and $C_d$ have been chosen,

In this section we describe how to condition the prior for transient displacements, $u$, with GNSS observations. We will assume that a appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_d$ have already been chosen.  Let $\mathbf{y}$ be the set of all $(x_i, t_j)$ pairs where a GNSS observation has been made, and let $\mathbf{d} = [d_{ij}]^T_{(x_i,t_j) \in \mathbf{y}}$. The random vector $\mathbf{d}$ is distributed as $\mathbf{d} \sim \mathcal{N}(\bar{\mathbf{d}},\mathbf{C_d})$, where 
\begin{equation}
\mathbf{d}
\end{equation}

We consider the column vector of GNSS observations, $\mathbf{d}_*$, to be realizations of $d_{ij}$ at the positions and times in $\mathbf{y}$ (i.e, $\mathbf{d}_* = [d_{ij}(\omega_*)]^T_{(x_i,t_j) \in \mathbf{y}}$). We want to find the posterior estimate of transient displacements, $u | \mathbf{d}_*$, which incorporates our prior and the GNSS observations.

Once we have chosen covariance functions and corresponding hyperparameters for $X$, $T$, and $C_d$, we then condition our prior Gaussian process, $u$, with the observations $d_*$ to form a posterior estimate of transient displacements, $u|d_*$. $d_*$ is a realization of the

Whichever the Gaussian processes describing transient displacements, $u$, and noise in the GNSS data, $\epsilon$, have covariance functions which
